---
title: "Reading VCFs and Building Mutational-Spectrum Catalogs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading VCFs and Building Mutational-Spectrum Catalogs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 8,
  fig.height = 3.5,
  dpi = 100,
  out.width = "100%"
)
have_bsgenome <- nzchar(system.file(package = "BSgenome.Hsapiens.1000genomes.hs37d5"))
have_msigplot <- nzchar(system.file(package = "mSigPlot"))
eval_all <- have_bsgenome && have_msigplot
knitr::opts_chunk$set(eval = eval_all)
```

This vignette walks through the four-step **mSigSpectra** pipeline —
`read_vcf` → `split_vcf` → `annotate_sbs_or_dbs_vcf` /
`annotate_id_vcf` → `vcf_to_catalog` — using a real Strelka VCF that
ships with the package. It then plots each catalog with
[`mSigPlot`](https://github.com/steverozen/mSigPlot).

If you do not have the BSgenome package or `mSigPlot` installed, the
chunks below will be skipped at render time. To install:

```r
BiocManager::install("BSgenome.Hsapiens.1000genomes.hs37d5")
remotes::install_github("steverozen/mSigPlot")
```

## 1. Load packages

```{r load}
library(mSigSpectra)
library(mSigPlot)
```

## 2. Locate test VCFs

Three example VCFs ship with the package: a Strelka SBS file, a Strelka
indel file, and a Mutect file (all small subsets, GRCh37 / hg19).

```{r files}
sbs_file <- system.file(
  "extdata", "Strelka-SBS-GRCh37", "Strelka.SBS.GRCh37.s1.vcf",
  package = "mSigSpectra"
)
id_file <- system.file(
  "extdata", "Strelka-ID-GRCh37", "Strelka.ID.GRCh37.s1.vcf",
  package = "mSigSpectra"
)
mutect_file <- system.file(
  "extdata", "Mutect-GRCh37", "Mutect.GRCh37.s1.vcf",
  package = "mSigSpectra"
)
```

## 3. Read the SBS VCF

`read_vcf()` returns a `data.table` with whatever columns the VCF body
contains. We pass `filter = "PASS"` to match the Strelka convention.

```{r read-sbs}
sbs_vcf <- read_vcf(sbs_file, filter = "PASS")
nrow(sbs_vcf)
head(sbs_vcf[, c("CHROM", "POS", "REF", "ALT", "FILTER")])
```

## 4. Split into SBS / DBS / ID sub-tables

`split_vcf()` partitions rows by `REF`/`ALT` length alone. For a
"pure SBS" VCF you would expect everything in `$SBS`, but Strelka's SBS
caller can emit adjacent SBS pairs that the indel caller does not see;
mSigSpectra **does not** merge them into DBSs (see "Gotchas" in the
README) — they stay as SBSs.

```{r split-sbs}
sbs_split <- split_vcf(sbs_vcf, name_of_vcf = "Strelka.SBS.GRCh37.s1")
sapply(sbs_split[c("SBS", "DBS", "ID")], nrow)
```

## 5. Annotate the SBS rows

`annotate_sbs_or_dbs_vcf()` adds:

* `seq.<N>bases` — the flanking sequence context (default `seq.21bases`).
* For ref genomes with a shipped transcript-ranges table (GRCh37 /
  GRCh38 / GRCm38), the transcript-strand columns
  `trans.start.pos`, `trans.end.pos`, `trans.strand`,
  `trans.Ensembl.gene.ID`, `trans.gene.symbol`, plus `bothstrand` and
  `count` (number of overlapping transcripts).

It returns a list with `annotated.vcf` (the table below) and
`discarded.variants` (always `NULL` for SBS / DBS).

```{r annotate-sbs}
sbs_ann <- annotate_sbs_or_dbs_vcf(sbs_split$SBS,
                                   ref_genome = "GRCh37")$annotated.vcf
new_cols <- setdiff(colnames(sbs_ann), colnames(sbs_split$SBS))
new_cols
```

## 6. Build SBS catalogs

A catalog is a single-column numeric matrix with attributes (no S3
class). One call per resolution.

```{r build-sbs}
cat96 <- vcf_to_catalog(sbs_ann, type = "SBS96", ref_genome = "GRCh37",
                        region = "genome", sample_name = "s1")
cat192 <- vcf_to_catalog(sbs_ann, type = "SBS192", ref_genome = "GRCh37",
                         region = "transcript", sample_name = "s1")
cat1536 <- vcf_to_catalog(sbs_ann, type = "SBS1536", ref_genome = "GRCh37",
                          region = "genome", sample_name = "s1")

dim(cat96)
sum(cat96)
catalog_attrs(cat96)
```

## 7. Plot SBS catalogs

```{r plot-sbs96, fig.height=3.5}
plot_SBS96(cat96, plot_title = "Strelka.SBS.GRCh37.s1 — SBS96")
```

```{r plot-sbs192, fig.height=4}
plot_SBS192(cat192, plot_title = "Strelka.SBS.GRCh37.s1 — SBS192 (stranded)")
```

```{r plot-sbs1536, fig.height=8}
plot_SBS1536(cat1536, plot_title = "Strelka.SBS.GRCh37.s1 — SBS1536")
```

## 8. Build and plot ID catalogs

The Strelka indel VCF goes through `annotate_id_vcf()`, which
left-justifies each indel, extracts the local repeat / microhomology
context, and emits the three classification strings (`COSMIC_83`,
`Koh_89`, `Koh_476`). Like `annotate_sbs_or_dbs_vcf()` it returns a
list of `annotated.vcf` + `discarded.variants`.

```{r id-pipeline}
id_vcf <- read_vcf(id_file, filter = "PASS")
id_split <- split_vcf(id_vcf, name_of_vcf = "Strelka.ID.GRCh37.s1")
id_ann <- annotate_id_vcf(id_split$ID,
                          ref_genome = "GRCh37")$annotated.vcf
id_ann[1, c("CHROM", "POS", "REF", "ALT", "COSMIC_83", "Koh_89")]

cat_id83 <- vcf_to_catalog(id_ann, type = "ID83", ref_genome = "GRCh37",
                           region = "genome", sample_name = "s1")
cat_id89 <- vcf_to_catalog(id_ann, type = "ID89", ref_genome = "GRCh37",
                           region = "genome", sample_name = "s1")
cat_id476 <- vcf_to_catalog(id_ann, type = "ID476", ref_genome = "GRCh37",
                            region = "genome", sample_name = "s1")
```

```{r plot-id83, fig.height=3.5}
plot_ID83(cat_id83, plot_title = "Strelka.ID.GRCh37.s1 — ID83")
```

```{r plot-id89, fig.height=4}
plot_ID89(cat_id89, plot_title = "Strelka.ID.GRCh37.s1 — ID89")
```

```{r plot-id476, fig.height=10}
plot_ID476(cat_id476, plot_title = "Strelka.ID.GRCh37.s1 — ID476")
```

## 9. Counts ↔ density

Convert from raw counts (per category) to mutation density (per
megabase of context) using the shipped k-mer abundances:

```{r density}
cat96_density <- transform_catalog(cat96,
                                   target_counts_or_density = "density")
attr(cat96_density, "counts_or_density")
```

```{r plot-sbs96-density, fig.height=3.5}
plot_SBS96(cat96_density,
           plot_title = "Strelka.SBS.GRCh37.s1 — SBS96 (density)")
```

## 10. Collapse from finer to coarser resolutions

```{r collapse}
cat96_from_1536 <- collapse_catalog(cat1536, to = "SBS96")
all.equal(as.numeric(cat96_from_1536[, 1]),
          as.numeric(cat96[, 1]))
```

## 11. Catalog I/O

ICAMS-native CSV is the default and only output format today;
SigProfiler and COSMIC formats are supported on input.

```{r io}
out_path <- file.path(tempdir(), "Strelka.SBS.GRCh37.s1.SBS96.csv")
write_catalog(cat96, out_path)
cat96_back <- read_catalog(out_path, region = "genome")
identical(as.numeric(cat96), as.numeric(cat96_back))
```

## See also

* `?read_vcf` — the caller-agnostic reader.
* `?annotate_sbs_or_dbs_vcf` — sequence context + transcript strand
  for SBS / DBS.
* `?annotate_id_vcf` — indel justification + classification (COSMIC 83
  / Koh 89 / Koh 476) + transcript strand.
* `?vcf_to_catalog` — single-call builder dispatching on `type`.
* `?transform_catalog` / `?collapse_catalog` — counts ↔ density and
  finer-to-coarser collapse.
* `?check_and_remove_discarded_variants` — optional defensive QC.
* The README's "Gotchas" section documents the cases this pipeline
  intentionally leaves to the user.
